How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in \textit{Emys marmorata} (Testudinoidea, Emydidae).

### 2015-01-16 doc/draft_turtle.tex
intro updated with ken comments

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
We collected landmark-based morphometric data from 354 adult \textit{E. marmorata} museum specimens. These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a} and represents adult individuals. We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. mamorata} and other emydines \citep{Angielczyk2013a}. We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives. When precise latitude and longitude information was not available we estimated it from whatever locality information was present. Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes. % How were adults chosen? Good to note how schemes differeed.

>>>>

% cryptic diversity
%   most species are still deliminated solely based on morphology
With the rise of molectular systematics and techniques such as DNA barcoding, it has been increasingly recognized that some taxa can only be differented using molecular data \citep{Bickford2007,SchlickSteiner2007,Stuart2006,Pfenninger2007,Funk2012,Clare2011}. Nevertheless, the majority of extant taxa, and nearly all extinct taxa, are delimited based solely via morphology. This disjuct is of great concern for studies of variation and diversity dynamics in deep time, where apparent morphological stasis may not reflect the true underlying diversity that existed \citep{Hunt2008,Eldredge1972,Gould1977a,VanBocxlaer2013}.

Much work has been devoted to species delimitation via sequence difference \citep{Fujita2012,Yang2010b} which in turn has sparked interest in whether geometric morphometric anlaysis can captur similarly fine-scale variation that can be used for identifying cryptic species. Most such studies focus on morphometrics to discover differences between taxa that were idetified by other means \citep{Polly2007a,Demandt2009,Gaubert2005b,Gunduz2007,Polly2003,Zelditch2004} and automated taxon identification \citep{MacLeod2007}. It would be very useful, though, if morphometrics approaches could reliably identify cryptic species without the use of other types of data. This would make the task of idenitfying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa.

Here, we attempt to address this issue using machine learning approaches. In particular, we ask whether it is possible to determine which amongst a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.

% past work on automatic taxon identification and older approaches to classifying taxa
% why use machine learning methods
\subsection{Background and system}
Machine learning is in many respects just an extension of known statistical methodology \citep{Hastie2009} in which the emphasis is placed on inferring rules and properties of data in order to explain the underlying structure. The basic statistical mechanics are supplemented by randomization, sorting, and partitioning algorithms and along with the maximization or minimization of summary statistics, in order to best estimate a general model for all data, both sampled and unsampled \citep{Hastie2009}. Machine learning approaches have found use in medical research, epidemiology, economics and automated image identification such as handwritten zip codes \citep{Hastie2009}. Two major classes of machine learning methods are unsupervised and supervised learning. Unsupervised learning methods are used with unlabeled data where the underlying structure is estimated and are analogous to clustering and density estimation methods \citep{Kaufman1990}. Supervised learning methods are used with labeled data where the final output of data is known and the rules for going from input to output are inferred. These are analogous classification and regression models \citep{Breiman1984}. The application of the alternative approaches used in this study illustrates only a sampling of the various previously derived methods for clustering observations and fitting classification models.

Geometric morphometric approaches to idetifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Zelditch2004,Mitteroecker2011,Polly2007a,Polly2003,Gunduz2007,Gaubert2005b,Demandt2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Francoy2009}. These methods are comparatively straightforward ways of understanding the differences in morphology between classes given their similarity to familiar multivariate approaches like Principal Components Analysis. They are benefit by producing results that can be easily visualized, which aids in the interpretation and presentation of data and results. Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. In this context, the study of \citet{Cardini2009a}, is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.

Here, we used multiple machine learning methods, both unsupervised and supervised, in order to compare different classification hypotheses. These methods provide different and unique advantages for understanding how to classify taxa, with what accuracy, and what these classifications are based on. While machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,VandenBrink2011,MacLeod2007}, including in the context of automated taxon identification and classification of groups, but the number of cases remains limited. In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.

\subsection{\textit{Emys marmorata}}
We applied our machine learning methods to the problem whether distinct subspecies or cryptic species extist within the western pond turtle, \textit{Emys marmorata} (formerly \emph{Clemmys marmorata}; see \citealp{Feldman2002}). \textit{E. marmorata} is distributed from northern Washington State, USA to Baja California, Mexico. Traditionally, \textit{E. marmorata} was classified into two named subspecies: the northern \textit{E. marmorata marmorata}, the southern \textit{E. marmorata palida} \citep{Seeliger1945}. She also recognized a central Californian intergrade zone between these subspecies, and she did not formally include the Baja California populations of \textit{E. marmorata} in these groups, implying the existence of a third distinct but unnamed subspecies. \textit{E. marmorata marmorata} is differentiated from \textit{E. marmorata palida} by the presence of a pair of triangular inguinal plates and darker neck markings. It should be noted that the triangular inguinal plates can sometimes be present in \textit{E. marmorata palida} though they are considerably smaller.

Previous work on morphological variation in \textit{E. marmorata} has focused primarily on differentiation between populations over a portion of the species' total range \citep{Lubcke2007,Germano2009,Germano2008,Bury2010}; comparatively few studies have included specimens from across the entire range \citep{Holland1992}. Most of these studies considered how local biotic and abiotic factors may contribute to differences in carapace length and found that size can vary greatly between different populations \citep{Lubcke2007,Germano2009,Germano2008}. There also has been interest in size-based sexual dimorphosim in \textit{E. marmorata} \citep{Lubcke2007,Germano2009,Holland1992}, with males being on average larger than females based on total carapace length and other linear measurements. However, the quality of size as a classifier of sex can vary greatly between populations \citep{Holland1992}, which makes sense in light of the amount of between population size difference \citep{Lubcke2007,Germano2009}. However, the effect of sexual dimorphism on shape, \textit{sensu} \citet{Kendall1977a}, was not assessed \citep{Holland1992,Lubcke2007,Germano2008}.

Of particular importance in the context of cryptic diversity in \textit{E. marmorata} is the traditional morphometric analysis of carapace shape carried out by \citet{Holland1992}. He compared morphological differences between and among many populations of \textit{E. marmorata} in three different areas of the species' range, and studied the relative effect of distance versus barriers in fostering morphological differentiation in the species. \citet{Holland1992} concluded that distance was a poor indicator of mophological differentiation, and the geographic features such as breaks between different drainage basis are probably more important barriers to reproduction. Additionally, \citet{Holland1992} found that morphological differences were observable as the magnitude of barriers and distance increased, but the variation required many variables to adequately capture, implying only very subtle morphological differentiation between putatively distinct populations. \citet{Holland1992} concluded that \textit{E. marmorata} is best classified as three distinct species: a northern species, souther species, and a Columbia basic species. This classification is similar to that of \citet{Seeliger1945}, except elevated to the species level and without recognition of a distinct Baja species.

More recently, the phylogeography of \textit{E. marmorata} and the possibility of cryptic diversity has been investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}. Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E. marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade. Analyses with nuclear DNA \citep{Spinks2010} with single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E. marmorata}, although the dataset differe in the location of this break point. In these papers, Spinks and colleagues discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E. marmorata} and \emph{E. pallida}), and a possible third based on populations in Baja California. However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}. Given that these characters are somewhat variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, we wondered whether a geometric morphometric analysis of shell shape would provide a more reliable way of diagnosing groups (whether species or subspecies) within \textit{E. marmorata}.

In this study, we attempt to estimate the best classification scheme of \textit{E. marmorata} based on variation in plastral (central shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data. Because of unclear geographic boundaries between subgroups of \textit{E. marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification. We hypothesize that if morphological variation corresponds to class assignment, then it should be possible to determine the best classification hypothesis of \textit{E. marmorata} from amongst multiple candidate hypotheses. However, if morphological variation variation does not correspond to any classification hypothesis, then supervised learning model generalization performance will be poor and reflect how variation may not follow along with any of the candidate classification hypotheses.

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
We collected landmark-based morphometric data from 354 adult \textit{E. marmorata} museum specimens. These specimens are a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a} and represents adult individuals. We chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. mamorata} and other emydines \citep{Angielczyk2013a}. \uppercase{Peter note -- how was something classified as an adult?}

We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives. When precise latitude and longitude information was not available we estimated it from whatever locality information was present. Because the specimens sampled to obtain the genetic data used to define the subclades were not available for study, all specimen classifications were based solely on the geographic information, not explicit assignment in previous studies. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two different schemes; se we compared a total of six different schemes. The schemes differed based on where geographic boundaries were assigned. This changes how certain individuals were assigned two one of the groups within in hypothesis such as which of the three morphologically defined groups, which of the four mitochondrially defined groups, and so on. % How were adults chosen? Good to note how schemes differeed.
